### 2022-09-15 main/chapter_searches.tex
cleanup, prepare for next writing

Misrepresentations of \pvalues\ (and other frequentist statistics) are
widespread and widely reported~\cite{
schervish1996p,
Cohen1994TheEI,
Amrhein2017TheEI,
goodman2008dirty,
greenland2016no,
Bernstein2016princess,
wagenmakers2007practical
}
in areas of they are not banned~\cite{Trafimow2015ban}.
For example, the false presentation as a ``probability of compatibility'' with
a model~\cite{HIGG-2018-04} appears to be spreading as a good
meme~\cite{dawkins1989selfish} in High Energy Physics ~\cite{
HIGG-2017-09,
HIGG-2018-27,
HIGG-2018-51,
EXOT-2018-08,
HION-2018-19,
mastrandrea2019searches,
white2019search,
langford2021combination,
IceCube2013search,
IceCube2014searches,
gerasimov2021new
},%
\footnote{%
A draft of~\cite{HIGG-2018-51} stated that a \pvalue\ determined
``the probability that the background-only hypothesis is compatible with the
observed data''.
That text was removed in response to my internal comments, but
``probability of compatibility'' remains in its figure captions.%
}%
\footnote{%
Reference~\cite{HIGG-2018-57} accurately describes a \pvalue\ as a tail
probability, but also (repeatedly) states that a
``probability of compatibility\ldots\ corresponds to a \pvalue''.%
}%
and is plausibly explained by mutations from vaguely accurate statements like
``compatible with the SM prediction with a $P$ value of $0.10\%$''~\cite{
lhcb2021test
}
to precisely inaccurate statements like
``a probability of around $0.1\%$ that the data is compatible with the
Standard Model predictions''~\cite{cern2021test}.

@@

Early variants of $\cls$ derived it from an identity with posterior tail
probabilities in Poisson or Gaussian special cases~\cite{
Helene1983upper,
pdg1988,
read2000modified,
pdg2022ynf
}
with a contrived frequentist interpretation~\cite{zech1988cls}.
For use in general cases, it is more recently described as a
``modified frequentist'',
``approximate confidence in the signal-only hypothesis''~\cite{
read2000modified,
Read2002cls
}.
High-profile usage in Higgs boson searches at the
Large Electron–Positron Collider~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches,
lep2003search
}
appears to have propelled $\cls$ to prominence;
these early applications used data deficits to exclude their
background-plus signal models with the test statistic $t(n) = -n$, but
disliked how deficits led to small \pvalues\ for the background-only model,
with which they might have confusingly set negative limits on the signal.
Such \pvalues\ are illustrated in Figure~\ref{fig:searches_sb_n} (left)
for small data counts;
in contrast, $\cls$ can only favour the background-only model with its negative
values shown on the right of this figure.

Although the statement in Section~\ref{sec:searches_fisher} that \pvalue\
ratios are not ``not \emph{necessarily} misleading'',
Figure~\ref{fig:searches_sb_n} shows a bias from the fact that
no data can give a $\cls$ that favours the signal model;
$\log\cls$ is non-positive with the $-n$ test statistic, and indeed non-negative
for the opposite ordering $t(n) = n$.
The $\log$ likelihood ratio on the actual data does not feature this bias:
it favours background-only for deficits and background-plus-signal for
excesses and is indeed linear in $n$:
\begin{equation}
\log \frac{L(s + b)}{L(b)}
= -s + n\log(1 + \frac{s}{b})
.
\end{equation}
Since likelihood ratio is monotonic in $n$, both $n$ and $-n$ are optimal
test statistics by the criteria of the Neyman-Pearson lemma.

By using \pvalue\ ratios with Neyman-Pearson optimal test statistic, we have
constructed an analysis that can only ever favour one hypothesis:
we have built a propaganda machine that only ever speaks truth but censors any
information that supports alternative ideas.
Reducing data to inequalities $t \geq \tilde{t}$ implements censorship,
and the likelihood-ratio ordering guarantees the one-sided inferences.

How can one escape from this trap?
Likelihood ratios on the actual data escape immediately by bypassing the
censor.
Even inequalities need not be biased --- if against a predetermined threshold,
say $5$ events, then models can predict whether or not $n \leq 5$, and its
truth or falsehood in data can favour either side.%
\footnote{%
This reduction to a fixed inequality appears to be the condition for which a
\pvalue\ ratio is not misleading when interpreted as a likelihood ratio.%
}
Finally, a reader can fall back to not trusting the reporter ---
the statements `\atlas\ reports $n \leq 5$' and logical `$n \leq 5$' are not
identical, since one can infer from the former that in fact $n = 5$.
They know what the message says, but not whether it states the whole truth.

@@

probability of compatibility~\cite{
HIGG-2018-04,
HIGG-2018-27,
IceCube2013search,
lhcb2021test,
cern2021test
}

mix $p$-value with posterior tail~\cite{lhcb2017test}

\subsection{Searches}
\label{sec:searches_searches}
% histograms, CR VR SR, systematics
% histfactory

\TODO{minimize energy not max logf}
To model the consequences of these many factors of uncertainty, each factor is
assigned a real-valued parameter, which when varied changed the expected yields
in the histogram bins of the search.
In this way, we form a statistical model which is a function
\begin{equation}
H_{\!f}(\vec \theta) \to (\vec \mu, \log f)
\end{equation}
from the parameters $\theta$ to Poisson expectations $\vec\mu$ of the histogram
bins and a regularization term $\log f$ that is maximized in fitting.

@@

lhcb2021test article: compatible with the SM prediction with a P value of 0.10\%
cern:  probability of around 0.1\% that the data is compatible with the Standard Model predictions
https://home.cern/news/news/physics/intriguing-new-result-lhcb-experiment-cern

\TODO{Statistical and systematic uncertainties}

>>>>

Misrepresentations of \pvalues\ (and other frequentist statistics) are
widespread and widely reported~\cite{
schervish1996p,
Cohen1994TheEI,
Amrhein2017TheEI,
goodman2008dirty,
greenland2016no,
Bernstein2016princess,
wagenmakers2007practical
}
in fields where they are not banned~\cite{Trafimow2015ban}.
For example, the false presentation as a ``probability of compatibility'' with
a model~\cite{HIGG-2018-04} appears to be spreading as a good
meme~\cite{dawkins1989selfish} in High Energy Physics ~\cite{
HIGG-2017-09,
HIGG-2018-27,
HIGG-2018-51,
EXOT-2018-08,
HION-2018-19,
mastrandrea2019searches,
white2019search,
langford2021combination,
IceCube2013search,
IceCube2014searches,
gerasimov2021new
},%
\footnote{%
A draft of~\cite{HIGG-2018-51} stated that a \pvalue\ determined
``the probability that the background-only hypothesis is compatible with the
observed data''.
That text was removed in response to my internal comments, but
``probability of compatibility'' remains in its figure captions.%
}%
\footnote{%
Reference~\cite{HIGG-2018-57} accurately describes a \pvalue\ as a tail
probability, but also (repeatedly) states that a
``probability of compatibility\ldots\ corresponds to a \pvalue''.%
}%
and is plausibly explained by mutations from vaguely accurate statements like
``compatible with the SM prediction with a $P$ value of $0.10\%$''~\cite{
lhcb2021test
}
to precisely inaccurate statements like
``a probability of around $0.1\%$ that the data is compatible with the
Standard Model predictions''~\cite{cern2021test}.

@@

Early variants of $\cls$ derived it from an identity with posterior tail
probabilities in Poisson or Gaussian special cases~\cite{
Helene1983upper,
pdg1988,
read2000modified,
pdg2022ynf
}
with a contrived frequentist interpretation~\cite{zech1988cls}.
For use in general cases, it is more recently described as a
``modified frequentist'',
``approximate confidence in the signal-only hypothesis''~\cite{
read2000modified,
Read2002cls
}.
High-profile usage in Higgs boson searches at the
Large Electron–Positron Collider~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches,
lep2003search
}
appears to have propelled $\cls$ to prominence;
these early applications used data deficits to exclude their
background-plus signal models with the test statistic $t(n) = -n$, but
disliked how deficits led to small \pvalues\ for the background-only model,
with which they might have confusingly set negative limits on the Higgs boson
signal cross-section.
Such \pvalues\ are illustrated in Figure~\ref{fig:searches_sb_n} (left)
for small data counts;
in contrast, $\cls$ can only favour the background-only model, due to its
negative values that are shown on the right of the same figure.

Although the statement in Section~\ref{sec:searches_fisher} that \pvalue\
ratios are not ``not \emph{necessarily} misleading'',
Figure~\ref{fig:searches_sb_n} shows a bias from the fact that
no data can give a $\cls$ that favours the signal model;
$\log\cls$ is non-positive with the $-n$ test statistic, and indeed non-negative
for the opposite ordering $t(n) = n$.
The $\log$ likelihood ratio on the actual data does not feature this bias:
it favours background-only for deficits and background-plus-signal for
excesses and is indeed linear in $n$:
\begin{equation}
\log \frac{L(s + b)}{L(b)}
= -s + n\log(1 + \frac{s}{b})
.
\end{equation}
Since the likelihood ratio is linear in $n$, both $n$ and $-n$ are optimal
test statistics by the criteria of the Neyman-Pearson lemma.

By using \pvalue\ ratios with Neyman-Pearson optimal test statistic, we have
constructed an analysis that can only ever favour one hypothesis:
we have built a propaganda machine that only ever speaks truth but censors any
information that supports alternative ideas.
Reducing data to inequalities $t \geq \tilde{t}$ implements the censorship,
and the likelihood-ratio ordering guarantees the one-sided inferences
that appear in Figure~\ref{fig:searches_sb_n} (right) as one-sided
$\log\cls$ values.

How can one escape this propaganda trap?
Likelihood ratios on the actual data escape immediately by bypassing the
censor.
Even inequalities need not be biased --- if taken against a predetermined
threshold, say $5$ events, then models can predict whether or not $n \leq 5$,
and its truth or falsehood in data can favour either side.%
\footnote{%
This reduction to a fixed inequality appears to be the condition for which a
\pvalue\ ratio is not misleading when interpreted as a likelihood ratio.%
}
Finally, a reader can fall back to not trusting the reporter ---
the statements `\atlas\ reports $n \leq 5$' and logical `$n \leq 5$' are not
identical, since one can infer from the former that in fact $n = 5$.
They know what the message says, but not that it states the whole truth.

@@

\TODO{histfactory}
\TODO{minimize energy not max logf}

To model the consequences of these many factors of uncertainty, each factor is
assigned a real-valued parameter, which when varied changed the expected yields
in the histogram bins of the search.
In this way, we form a statistical model which is a function
\begin{equation}
H_{\!f}(\vec \theta) \to (\vec \mu, \log f)
\end{equation}
from the parameters $\theta$ to Poisson expectations $\vec\mu$ of the histogram
bins and a regularization term $\log f$ that is maximized in fitting.

@@

\TODO{uncertainties: systematics, statistical}

\subsection{\atlas\ SUSY searches}
\label{sec:searches_searches}

\TODO{CR VR SR}
